---
title: "Total Nitrogen"
author: "L. Patterson and J. Fay"
date: "Spring, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<style type="text/css">

body{ /* Normal  */
      font-size: 18px;
  }
</style>

# Exploratory Data Analysis

## Prep the workspace by installing (if needed) and loading libraries
The method for installing and loading libraries, as well as downloading data from the USGS, are explained in the `LoadStreamflowDescription.Rmd` file. `LoadStreamflowDescriptions.Rmd` file. 

```{r libinstall, message=FALSE, warning=FALSE, include=FALSE}
#Import 'dataRetrieval' library (to scrape USGS data)
if (!requireNamespace("dataRetrieval", quietly = TRUE))
  install.packages("dataRetrieval")
require(dataRetrieval)

#Import 'EGRET' library (for river data analysis and vis)
if (!requireNamespace("EGRET", quietly = TRUE))
  install.packages("EGRET")
require(EGRET)

#Import 'ggplot2' library (for plotting)
if (!requireNamespace("ggplot2", quietly = TRUE))
  install.packages("ggplot2")
require(ggplot2)

#Import 'trend' library (for trend modeling)
if (!requireNamespace("trend", quietly = TRUE))
  install.packages("trend")
require(trend)

#Import 'lubridate' library (for easy date manipulation)
if (!requireNamespace("lubridate", quietly = TRUE))
  install.packages("lubridate")
require(lubridate)

#Import 'dplyr' library (for data manipulation)
if (!requireNamespace("dplyr", quietly = TRUE))
  install.packages("dplyr")
require(dplyr)

#Import 'magrittr' library (for using pipes in code)
if (!requireNamespace("magrittr", quietly = TRUE))
  install.packages("magrittr")
require(magrittr)

#Import 'leaflet' library (for mapping)
if (!requireNamespace("leaflet", quietly = TRUE))
  install.packages("leaflet")
require(leaflet)
```

```{r libs, message=FALSE, warning=FALSE}
library(dataRetrieval); library(EGRET)
library (ggplot2);  library(trend); library(lubridate)
library(dplyr); library(magrittr); library(leaflet)
```
<br />

## Load in the water quality *site* data
(The data should be saved in the data folder one up from the current R script...)
```{r downloaddata, message=FALSE, warning=FALSE}
#set working directory
swd <- "../data/"

#read in the csv files  
sites <- read.csv(paste0(swd,"station.csv"), sep=',',header=TRUE); dim(sites)
  sites[1:3,]
  dim(sites)
```
<br />

## Plot the sites

Load in the site data and plot a map to see where water quality data were present from this portal. 


Let's play with the data and see what it looks like to subset the data by removing small streams with a drainage area of less than 25 mi2.
```{r plotSites}
#plot sites
map <-  leaflet(data=sites) %>% 
    addProviderTiles("CartoDB.Positron") %>%
    addCircleMarkers(~LongitudeMeasure,~LatitudeMeasure,
                     color = "red", radius=3, stroke=FALSE,
                     fillOpacity = 0.8, opacity = 0.8,
                     popup=~MonitoringLocationName)  
map
```

<br />

When we do this, we find there are only 10 sites remaining. This seems really small, so we plot these new sites on the map (blue), overlaying the orginal sites in red.

```{r sites2}
# subset to only include those with a drainage area exceeding 25 square miles
sites2 <- subset(sites, DrainageAreaMeasure.MeasureValue >= 25); dim(sites2)

#where are sites located?
leaflet(data=sites) %>%  
    addProviderTiles("CartoDB.Positron") %>%
    addCircleMarkers(~LongitudeMeasure,~LatitudeMeasure,
                   color = "red", radius=3, stroke=FALSE,
                   fillOpacity = 0.8, opacity = 0.8,
                   popup=~MonitoringLocationName) %>%
    addCircleMarkers(~sites2$LongitudeMeasure,~sites2$LatitudeMeasure,
                 color = "blue", radius=3, stroke=FALSE,
                 fillOpacity = 0.8, opacity = 0.8,
                 popup=~MonitoringLocationName)
#Notice for Jordan Lake... the two values we will be most interested in are: Haw River at SR 1943 and 
#All the names on Jordan Lake are "Jordan Lake"... change popup name to see options
```

<br />

We take a closer look at the drainage measurement values and see there are numerous NA values. Clearly, filtering by drainage area will not look. The main point I want to make here is **don't be afraid to *play* with the data**.

Let's replot the site map but now change the popup name to see which site names are located within each branch of Jordan Lake.

```{r changePopup}
leaflet(data=sites) %>%  
  addProviderTiles("CartoDB.Positron") %>%
  addCircleMarkers(~LongitudeMeasure,~LatitudeMeasure,
                   color = "red", radius=3, stroke=FALSE,
                   fillOpacity = 0.8, opacity = 0.8,
                   popup=~MonitoringLocationIdentifier)

#reduce columns to those needed
sites <- sites[,c(1,3:4,8:9,12:13)]
```


## Load in the water quality *measurement* data

Once we load in the measurements data, we need to do some cleaning. The first step is to *filter* the data to what we want to use. For example, 

* We want to make sure we are using routine samples and **not extreme event sampling** (biased for specific occasions and not for estimating annual average load).<br>

* We specify what **type of nitrogen** we want to use. From the literature we found that regulations for nitrogen include: nitrate, nitrite, ammonia and organic forms. Doing some reading about the WQX standards, you learn that nitrogen, mixed froms incorporates all of the above forms of nitrogen.<br>

* We also want to make sure we are looking at **total Nitrogen**, so make sure the Results Sample Fraction Text only includes those with Total.

A filtered dataset is provided for you as the `results.csv` in the data folder. We'll load this in and resume analysis...
```{r downloadresultsdata, message=FALSE, warning=FALSE}
results <- read.csv(paste0(swd,"result.csv"), sep=',',header=TRUE); dim(results)
  names(results)

#find out what options are present for the Hydrologic Events and Characteristic Name
unique(results$HydrologicEvent);
unique(results$CharacteristicName);

#filter data
nitrogen <-results %>% 
          filter(ActivityMediaSubdivisionName=="Surface Water") %>% 
          filter(HydrologicEvent != "Storm" & HydrologicEvent != "Flood" & HydrologicEvent != "Spring breakup" & HydrologicEvent != "Drought") %>%
          filter(CharacteristicName=="Nitrogen, mixed forms (NH3), (NH4), organic, (NO2) and (NO3)") %>%
          filter(ResultSampleFractionText=="Total")
dim(nitrogen)

#check that filters worked
unique(nitrogen$HydrologicEvent); unique(nitrogen$ActivityMediaSubdivisionName)
unique(nitrogen$CharacteristicName);  unique(nitrogen$ResultSampleFractionText)

#reduce to needed columns
nitrogen <- nitrogen[,c(1,7,22,31:35,59:61)]
```

### Detection Limits and Unit Conversion

You may have noticed that many sample sites state "not detected". This is important data that are not currently being represented. Create a new column and set the value equal to the results, unless it is below the detection limit - in which case set it equal to ½ of the detection limit.

You may also have noted that the total nitrogen was sometimes reported as mg/l or mg/l NO3. We want mg/l. To convert to mg/l, we know the atomic weight of nitrogen is 14.0067 and the molar mass of nitrate anion (NO3) is 62.0049 g/mole. Therefore, to convert between units:

* Nitrate-N (mg/L) = 0.2259 x Nitrate-NO3 (mg/L)
* Nitrate-NO3 (mg/L) = 4.4268 x Nitrate-N (mg/L) 


``` {r detection}
#set data below detection limits equal to 1/2 the detection limit
nitrogen$TotalN <- ifelse(nitrogen$ResultDetectionConditionText=="Not Detected", nitrogen$DetectionQuantitationLimitMeasure.MeasureValue/2, nitrogen$ResultMeasureValue)

#convert mg/l as No3 to mg/l as N
nitrogen$TotalN <- ifelse(nitrogen$ResultMeasure.MeasureUnitCode=="mg/l NO3", nitrogen$TotalN*0.2259, nitrogen$TotalN)
nitrogen$TotalN <- ifelse(nitrogen$DetectionQuantitationLimitMeasure.MeasureUnitCode=="mg/l NO3", nitrogen$TotalN*0.2259, nitrogen$TotalN)
```


## Exploratory Data Analysis
Now that we have explored and tidied our `sites` and `results` datasets pulled from the Water Quality Data portal, we'll **merge** them together and generate our visualizations. Recalling our objective is to reveal trends in water quality, particularly in repsonse to water quality rules issued in 2009, we'll want to constrain (i.e. **filter**) our analyses to sites with data collected before and after 2009. It's these remaining sites that we'll construct **plots** of total nitrogen: line plots of N over time, box plots by year, and box plots by month. We'll also plot our concentrations on a **map**. 

Additionally, we'll convert our N concentrations to annual load (lbs/year), which will require pulling in discharge data (as we did in previous sessions). We'll do this for a few sites, comparing the findings to allowable levels to answer: ***are these sites in compliance with Total Daily Maximum Loads (TMDLs)?***

### Merge sites and measurement data together
Currently, site and measurement data are not connected together. However, we may want to show the nitrate values on a map. To do this, we merge the data together based on a unique identifier shared between the two data sets. In this case, it is the *MonitoringLocationIdentifier* column.

```{r merge}
nitrogen.data <- merge(sites, nitrogen, by.x="MonitoringLocationIdentifier", by.y="MonitoringLocationIdentifier"); dim(nitrogen.data)

#create year and month categories. Make a table to see how many samples were collected each year/month
  nitrogen.data$Year <- year(nitrogen.data$ActivityStartDate);                    table(nitrogen.data$Year)
  nitrogen.data$Month <- month(nitrogen.data$ActivityStartDate);                  table(nitrogen.data$Month)
```

### Filter merged data

With the data merged, we now want to know how many sites are collecting data of interest. We find there are 22 sites. Of those sites, we want to know which were collecting data after the Jordan Lake Rules were passed and how much data are being collected at this site.

Based on this, we see several sites stopped collecting data prior to 2009, when the first iteration of Jordan Rules were passed. We also see that some of these sites collected only a few years of data. Remove those sites and plot the remaining sites on the map. 

```{r filterSitesMore}
#ED to know which sites to keep or remove
unique.sites <- unique(nitrogen.data$MonitoringLocationIdentifier)
  n.sites <- length(unique.sites)

#Create dataframe of information on: n samples, start year, end year
site.info <- nitrogen.data %>%
  group_by(MonitoringLocationIdentifier) %>%
  summarise(n=n(), StartYear=min(Year), EndYear=max(Year))
site.info <- as.data.frame(site.info) 
site.info$Range <- site.info$EndYear-site.info$StartYear
site.info

#Limit data to those with more than 10 years data and still operating
site.info <- filter(site.info, Range>=10 & EndYear>=2017); site.info
df <- nitrogen.data[nitrogen.data$MonitoringLocationIdentifier %in% unique(site.info$MonitoringLocationIdentifier), ]; dim(df)

#Where are they located
leaflet(data=df) %>%  addProviderTiles("CartoDB.Positron") %>%
  addCircleMarkers(~LongitudeMeasure,~LatitudeMeasure,
                   color = "red", radius=5, stroke=FALSE,
                   fillOpacity = 0.8, opacity = 0.8,
                   popup=~MonitoringLocationIdentifier)
```

###Plot total nitrogen over time, by month and by year for each site.

```{r NoverTime, message=FALSE, warning=FALSE}
####Plot nitrogen over time at each site
unique.site <- unique(df$MonitoringLocationIdentifier)

#convert activity date into a data
df$Date <- as.Date(df$ActivityStartDate, "%Y-%m-%d")
df$Year <- year(df$Date);   df$Month <- month(df$Date)


i=1
for (i in 1:length(unique.site)){
  zt <- filter(df, MonitoringLocationIdentifier == unique.site[i])
  
  #group by date - take average of a day
  foo <- zt %>% group_by(Date) %>% 
    summarise(AverageN = mean(TotalN, na.rm=TRUE), Year=mean(Year), Month=mean(Month))
  foo <- as.data.frame(foo)
  
  foo$color <- rgb(0.596,0.961,1,0.5)
  foo$color <- ifelse(foo$Month==1, rgb(0.478,0.772,0.804,0.8), foo$color)
  foo$color <- ifelse(foo$Month==2, rgb(0.325,0.525,0.545,0.8), foo$color)  
  foo$color <- ifelse(foo$Month==3, rgb(0.792,1,0.439,0.8), foo$color)    
  foo$color <- ifelse(foo$Month==4, rgb(0.635,0.804,0.352,0.8), foo$color)    
  foo$color <- ifelse(foo$Month==5, rgb(0.333,0.420,0.184,0.8), foo$color)  
  foo$color <- ifelse(foo$Month==6, rgb(1,0.447,0.337,0.8), foo$color)
  foo$color <- ifelse(foo$Month==7, rgb(0.804,0.357,0.114,0.8), foo$color)  
  foo$color <- ifelse(foo$Month==8, rgb(0.545,0,0,0.8), foo$color)  
  foo$color <- ifelse(foo$Month==9, rgb(1,0.765,0.059,0.8), foo$color)  
  foo$color <- ifelse(foo$Month==10, rgb(0.804,0.584,0.047,0.8), foo$color)  
  foo$color <- ifelse(foo$Month==11, rgb(0.545,0.396,0.031,0.8), foo$color)  
  
  #create a plot
  par(mfrow=c(1,1))
  layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE), widths=c(2,2), heights=c(1.5,1.5))
  par(mar = c(2,5,2,2)) #set plot margins
  
  #Plot single graph on top row
  plot(foo$Date, foo$AverageN, type='n', yaxt="n", main=zt$MonitoringLocationIdentifier[i], ylab="Total Nitrogen (mg/l)", xlab = '', cex.lab=0.9, ylim=c(0,6))
    axis(2, las=2, cex.axis=0.8)
  lines(foo$Date, foo$AverageN, col=rgb(0,0,0,0.4))  
  points(foo$Date, foo$AverageN, col=foo$color, cex=0.8, pch=19)  
  
  legend_order <- matrix(1:12,ncol=4,byrow = TRUE)
  legend("topright", c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"),
         pch=19, col=c(rgb(0.478,0.772,0.804,0.8),rgb(0.325,0.525,0.545,0.8),rgb(0.792,1,0.439,0.8),rgb(0.635,0.804,0.352,0.8),rgb(0.333,0.420,0.184,0.8),
                       rgb(1,0.447,0.337,0.8),rgb(0.804,0.357,0.114,0.8),rgb(0.545,0,0,0.8),rgb(1,0.765,0.059,0.8),rgb(0.804,0.584,0.047,0.8),
                       rgb(0.545,0.396,0.031,0.8), rgb(0.596,0.961,1,0.5)), ncol=4)

#Add blank months to foo in case
  blank.months <- as.data.frame(matrix(nrow=12,ncol=5)); colnames(blank.months) <- colnames(foo)
  months.tab <- as.data.frame(table(foo$Month)); colnames(months.tab)<-c("Month","Freq");
  k=1
  for (j in 1:12){
    check <- j; 
    check.zt <- subset(months.tab, Month==j)
  
    if(dim(check.zt)[1]==0){
      blank.months$Month[k]=check;   blank.months$AverageN[k]=0
      k=k+1    
    }
  }
  blank.months <- blank.months[!is.na(blank.months$Month),]
  blank.months
  foo.mo <- rbind(foo, blank.months)
  
  boxplot(foo.mo$AverageN~foo.mo$Month, yaxt="n", ylab="Total Nitrogen (mg/l)", ylim=c(0,10), yaxs='i',
          col=c(rgb(0.478,0.772,0.804,0.8),rgb(0.325,0.525,0.545,0.8),rgb(0.792,1,0.439,0.8),rgb(0.635,0.804,0.352,0.8),rgb(0.333,0.420,0.184,0.8),
                rgb(1,0.447,0.337,0.8),rgb(0.804,0.357,0.114,0.8),rgb(0.545,0,0,0.8),rgb(1,0.765,0.059,0.8),rgb(0.804,0.584,0.047,0.8)), pch=19, cex=0.7)
  axis(2, las=2)
  mtext("Nitrogen by Month", side=3, line=-2)
  abline(h=0.01, col="black", lwd=1)
  
  
  #Add blank years to foo in case
  year.tab <- as.data.frame(table(foo$Year)); colnames(year.tab)<-c("Year","Freq");
  year.tab$Year <- as.numeric(as.character(year.tab$Year))
  row.no <- max(year.tab$Year)-min(year.tab$Year)
  
  blank.years <- as.data.frame(matrix(nrow=row.no,ncol=5)); colnames(blank.years) <- colnames(foo)
  
  k=1
  
  for (j in min(year.tab$Year):max(year.tab$Year)){
    check <- j; 
    check.zt <- subset(year.tab, Year==check)
    
    if(dim(check.zt)[1]==0){
      blank.years$Year[k]=check;   blank.years$AverageN[k]=0
      k=k+1    
    }
  #print(check)
  }
  blank.years <- blank.years[!is.na(blank.years$Year),]
  blank.years
  
  foo.yr <- rbind(foo, blank.years)
  
  boxplot(foo.yr$AverageN~foo.yr$Year, yaxt="n", ylab="Total Nitrogen (mg/l)", ylim=c(0,6), col="grey", pch=19, cex=0.7)
  axis(2, las=2)
  mtext("Nitrogen by Year", side=3, line=-2)
  abline(h=10, col="red", lwd=2, lty=4)
  
  #print(i)
}

```

### Plot the 2017 Nitrogen year on leaflet

Let's show the amount of Nitrogen coming from each site based on the size of the circle.

``` {r sizeLeaflet}
nitrogen.data$Year <- year(nitrogen.data$ActivityStartDate)
last.year <- filter(nitrogen.data, Year==2017)
last.year.n <- last.year %>%
  group_by(MonitoringLocationIdentifier) %>%
  summarize(AveN = mean(TotalN, na.rm=T), LongitudeMeasure = mean(LongitudeMeasure), LatitudeMeasure = mean(LatitudeMeasure))
last.year.n <- as.data.frame(last.year.n)

last.year.n$AveN <- round(last.year.n$AveN,3)

leaflet(data=last.year.n) %>%  
  addProviderTiles("Esri.WorldTopoMap") %>%
  addCircleMarkers(~LongitudeMeasure,~LatitudeMeasure,
                   color = "red", radius=~AveN*5, stroke=FALSE,
                   fillOpacity = 0.6, opacity = 0.8,
                   popup=~paste0(MonitoringLocationIdentifier, "<br>",
                          "Average 2017 N: ", AveN, " mg/l"))
```

<br />

## Compare the Nitrogen Load with Thresholds

The water quality data reports Nitrogen as mg/l. In order to convert to an annual load (lbs/yr), we need to know the volume of water flowing through each site. Go to the NWIS Mapper to find which USGS gauges are closest to the Haw River Arm (1 site) and the New Hope Arm (3 sites).

### Load NWIS data for Haw River
``` {r NWISHaw}
#USGS-0209699999 is the arm of the Haw River and there is a USGS stream gauge site nearby: USGS 02096960
#Identify gauge to download
siteNo <- '02096960' 
#Identify parameter of interest: https://help.waterdata.usgs.gov/code/parameter_cd_query?fmt=rdb&inline=true&group_cd=%
pcode = '00060' #discharge (cfs)
#Identify statistic code for daily values: https://help.waterdata.usgs.gov/code/stat_cd_nm_query?stat_nm_cd=%25&fmt=html
scode = "00003"  #mean
#Identify start and end dates
start.date = "1970-10-01"
end.date = "2017-12-31"

#Load in data using the site ID
q <- readNWISdv(siteNumbers = siteNo, parameterCd = pcode, statCd = scode, startDate=start.date, endDate=end.date)
  q <- renameNWISColumns(q); colnames(q)
q.siteInfo <- attr(q, "siteInfo")
q$Year <- year(q$Date);  q$Month <- month(q$Date)

haw.n <- filter(df, MonitoringLocationIdentifier == "USGS-0209699999")

#where are sites located?
leaflet(data=q.siteInfo) %>%  
  addProviderTiles("Esri.WorldTopoMap") %>%
  addCircleMarkers(~dec_lon_va,~dec_lat_va,
                   color = "blue", radius=3, stroke=FALSE,
                   fillOpacity = 0.8, opacity = 0.8,
                   popup=~station_nm) %>%
  addCircleMarkers(~haw.n$LongitudeMeasure[1],~haw.n$LatitudeMeasure[2],
                   color = "red", radius=3, stroke=FALSE,
                   fillOpacity = 0.8, opacity = 0.8,
                   popup=~haw.n$MonitoringLocationName[1])
```

### Calculate Annual Load for Haw River

To calculate the annual load we need to convert cfs to MGD. Then we use pipes and dplyr to calculate the total annual flow at Haw River. Next, we calculate the average Nitrogen load for samples taken during each year. This is a rough proxy. A finer analysis can be undertaken by summarizing monthly flow and water quality, aggregating to the year as the last step.
The annual load is then the: **Total Flow * average Nitrogen * 8.34 lbs per gallon**

We can then plot the annual load with the threshold of 2.567 Million pounds per year.

```{r HawLoad}
#Calculate total volume of water each year
q$Flow_MGD <- 646317*q$Flow/1000000
q.year <- q %>% 
  group_by(Year) %>%  
  summarise(Total_MGD = sum(Flow_MGD, na.rm=T), n=n())
q.year <- as.data.frame(q.year)
q.year <- filter(q.year, n>=350)


#calculate the average Nitrate concentration
n.year <- haw.n %>% 
  group_by(Year) %>%  
  summarise(AveN = mean(TotalN, na.rm=T), n=n())
n.year <- as.data.frame(n.year)



#For days with a water quality measurement - what was the daily flow?
haw.n <- merge(n.year, q.year, by.x="Year", by.y="Year")
lbspergal <- 8.34
haw.n$Pounds <- haw.n$Total_MGD*haw.n$AveN*lbspergal
#convert to million pounds
haw.n$MPounds <- haw.n$Pounds/1000000

par(mfrow=c(1,1))
par(mar = c(2,5,2,2)) #set plot margins

#Plot single graph on top row. 
plot(haw.n$Year, haw.n$MPounds, type='n', yaxt="n", main="Haw River Branch", ylab="Annual Nitrogen Load (million lbs)", xlab = '', cex.lab=0.9)
  axis(2, las=2, cex.axis=0.8)
  lines(haw.n$Year, haw.n$MPounds, col=rgb(0,0,0,0.4))  
  points(haw.n$Year, haw.n$MPounds, col=rgb(0,0,0,0.6), cex=1, pch=19)  
abline(h=2567000/1000000, col="red", lty=4, lwd=2)

```

### Calculate Annual Load for New Hope Creek
Here there are 3 upstream gauges for New Hope Creek. We will download that information into a single file. We repeat the above analyis with the New Hope stream gauges. 

```{r NewHopeLoad}
#USGS-0209768310 is the arm of the Haw River and there are three usgs gauges upstream on different tributaries
#Identify gauges to download
siteNo <- c('02097314', '02097517','0209741955')
pcode = '00060' #discharge (cfs)
scode = "00003"  #mean
start.date = "1970-10-01"
end.date = "2017-12-31"

#Load in data using the site ID
q <- readNWISdv(siteNumbers = siteNo, parameterCd = pcode, statCd = scode, startDate=start.date, endDate=end.date)
q <- renameNWISColumns(q); colnames(q)
q.siteInfo <- attr(q, "siteInfo")
q$Year <- year(q$Date);  q$Month <- month(q$Date)
#If flow is unknown set to NA
q$Flow <- ifelse(q$Flow <0,NA, q$Flow)

#grab new hope monitoring station
newhope.n <- filter(df, MonitoringLocationIdentifier == "USGS-0209768310")

#where are sites located?
leaflet(data=q.siteInfo) %>%  
  addProviderTiles("Esri.WorldTopoMap") %>%
  addCircleMarkers(~dec_lon_va,~dec_lat_va,
                   color = "blue", radius=3, stroke=FALSE,
                   fillOpacity = 0.8, opacity = 0.8,
                   popup=~station_nm) %>%
  addCircleMarkers(~newhope.n$LongitudeMeasure[1],~newhope.n$LatitudeMeasure[2],
                   color = "red", radius=3, stroke=FALSE,
                   fillOpacity = 0.8, opacity = 0.8,
                   popup=~newhope.n$MonitoringLocationName[1])

#Calculate total volume of water each year
q$Flow_MGD <- 646317*q$Flow/1000000
q.year <- q %>% 
  group_by(Year) %>%  
  summarise(Total_MGD = sum(Flow_MGD, na.rm=T), n=n())
q.year <- as.data.frame(q.year)
q.year <- filter(q.year, n>=350*3)


#calculate the average Nitrate concentration
n.year <- newhope.n %>% 
  group_by(Year) %>%  
  summarise(AveN = mean(TotalN, na.rm=T), n=n())
n.year <- as.data.frame(n.year)

#For days with a water quality measurement - what was the daily flow?
newhope.n <- merge(n.year, q.year, by.x="Year", by.y="Year")
lbspergal <- 8.34
newhope.n$Pounds <- newhope.n$Total_MGD*newhope.n$AveN*lbspergal
newhope.n$KPounds <- newhope.n$Pounds/1000

par(mfrow=c(1,1))
par(mar = c(2,5,2,2)) #set plot margins

#Plot single graph on top row
plot(newhope.n$Year, newhope.n$KPounds, type='n', yaxt="n", main="New Hope Arm", ylab="Annual Nitrogen Load (thousand lbs)", xlab = '', cex.lab=0.9, ylim=c(0,800))
  axis(2, las=2, cex.axis=0.8)
  lines(newhope.n$Year, newhope.n$KPounds, col=rgb(0,0,0,0.4))  
  points(newhope.n$Year, newhope.n$KPounds, col=rgb(0,0,0,0.6), cex=1, pch=19)  
abline(h=641021/1000, col="red", lty=4, lwd=2)
```

<br />

Notice that the annual nitrogen load is far below the threshold level. Why or why not is this unexpected? What do you think could be causing this underestimate?

It is likely due to missing several tributary contributions. Jordan Lake has a 1,689 mi2 drainage area. The Haw River Branch is 1,275 mi2. A tributary downstream of newhope.n is 12 mi2. The upstream drainage area of the three sites are 76.9, 21.1, and 41 mi2 (how do you think I found these numbers?). This leaves 402 mi2 unaccounted for. Let's assume 25% of this left overis downstream of the monitoring site. In hydrology, flow is often linked to drainage area. Let's adjust the Nitrogen entering the New Hope Arm by drainage area.

```{r adjustArea}
site.da <- 76.9+21.2+41
remaining.da <- 1689-1275-12
fraction <- 0.75

adjust.da <- remaining.da*fraction
newhope.n$adj.pounds <- newhope.n$KPounds*adjust.da/site.da

plot(newhope.n$Year, newhope.n$adj.pounds, type='n', yaxt="n", main="New Hope Arm", ylab="Annual Nitrogen Load (thousand lbs)", xlab = '', cex.lab=0.9, ylim=c(0,1000))
axis(2, las=2, cex.axis=0.8)
  lines(newhope.n$Year, newhope.n$KPounds, col=rgb(0,0,0,0.4))  
  points(newhope.n$Year, newhope.n$KPounds, col=rgb(0,0,0,0.6), cex=1, pch=19)  

  lines(newhope.n$Year, newhope.n$adj.pounds, col=rgb(0.5,0,0,0.4), lty=2)  
  points(newhope.n$Year, newhope.n$adj.pounds, col=rgb(0.5,0,0,0.4), cex=1, pch=19)  
abline(h=641021/1000, col="red", lty=4, lwd=2)
mtext(paste0("Adjust by ", fraction, " drainage area fraction"), side=3, line=-2)
```

##Save out the new files for entry into Tableau

```{r}
#write csv
write.csv(df, paste0(swd,'../tableau/nitrogen.csv'),row.names=FALSE)
write.csv(haw.n, paste0(swd, '../tableau/hawbranch.csv'),row.names=FALSE)
write.csv(newhope.n, paste0(swd, '../tableau/newhope.csv'),row.names=FALSE)
```